Visual Diagnostics for Constrained Optimisation with Application to Guided Tours

### 2021-02-18 paper/zhang-cook-laa-langrene-menendez.Rmd
better examples of projection pursuit indices

---
title: "Visual Diagnostics for Constrained Optimisation with Application to Guided Tours"
author:
  - name: H.Sherry Zhang
    affiliation: Monash University 
    address: Department of Econometrics and Business Statistics
    email:  huize.zhang@monash.edu
  - name: Dianne Cook
    affiliation: Monash University 
    address: Department of Econometrics and Business Statistics
    email:  dicook@monash.edu
  - name: Ursula Laa
    affiliation: University of Natural Resources and Life Sciences
    address: Institute of Statistics
    email:  ursula.laa@boku.ac.at  
  - name: Nicolas Langrené
    affiliation: CSIRO Data61 
    address: 34 Village Street, Docklands VIC 3008 Australia
    email: nicolas.langrene@csiro.au
  - name: Patricia Menéndez
    affiliation: Monash University 
    address: Department of Econometrics and Business Statistics
    email:  patricia.menendez@monash.edu 
abstract: |
  Guided tour searches for interesting low-dimensional views of high-dimensional data via optimising a projection pursuit index function. The first paper of projection pursuit by @friedman1974projection stated that "the technique used for maximising the projection index strongly influences both the statistical and the computational aspects of the procedure." While much work has been done in proposing  indices in the literature, less has been done on evaluating the performance of the optimisers. In this paper, we implement a data collection object in the optimisation of projeciton pursuit guided tour and introduce visual diagnostics based on the data object collected. These diagnostics and workflows can be applied to a broad class of optimisers, to assess their performance. An R package, \pkg{ferrn}, has been created to implement the diagnostics.

@@

```{r }
# spelling::spell_check_files(here::here("paper/zhang-cook-laa-langrene-menendez.Rmd"), lang = "en-GB", ignore = spelling::get_wordlist())
```

@@

Let $\mathbf{X}_{n \times p}$ be the data matrix with $n$ observations in $p$ dimensions. A $d$-dimensional projection is a linear transformation from $\mathbb{R}^p$ into $\mathbb{R}^d$ defined as $\mathbf{Y} = \mathbf{X} \cdot \mathbf{A}$, where $\mathbf{Y}_{n \times d}$ is the projected data and $\mathbf{A}_{p\times d}$ is the projection basis matrix. We define $f: \mathbb{R}^{n \times d} \mapsto \mathbb{R}$ to be an index function that maps the projected data $\mathbf{Y}$ onto a scalar value. This is commonly known as the projection pursuit index function, or just index function, and is used to measure the "interestingness" of a given projection. In the context of projection pursuit, interestingness means departure from normality [@diaconis1984asymptotics; @huber1985projection]. \textcolor{red}{only include a couple of indices, using e.g.}A number of index functions have been proposed in the literature to detect different data structures [@friedman1974projection; @hall1989polynomial; @cook1993projection; @posse1995projection] and to classify groups [@lee2005projection; @lee2010projection].

As a general visualisation method, tour produces animations of high dimensional data via rotations of low-dimensional planes. There are different tour implementations depending on how the high dimensional space is investigated: grand tour [@cook2008grand] selects the planes randomly to provide a general overview;  manual tour [@cook1997manual] gradually phases in and out one variable to understand the contribution of that variable in the projection. Guided tour, the main interest of this paper, chooses the planes with the aid of projection pursuit, to gradually reveal the most interesting projection. Given a random start, projection pursuit iteratively finds bases with higher index values and the guided tour constructs a geodesic interpolation between these planes to form a tour path. Figure \ref{fig:tour-path} shows a sketch of the tour path where the blue frames are produced by the projection pursuit optimisation and the white frames are interpolations  between the blue frames. Mathematical details of the geodesic interpolation can be found in @buja2005computational. The aforementioned tour method has been implemented in the R package \CRANpkg{tourr} [@tourr].

>>>>

---
title: "Visual Diagnostics for Constrained Optimisation with Application to Guided Tours"
author:
  - name: H.Sherry Zhang
    affiliation: Monash University 
    address: Department of Econometrics and Business Statistics
    email:  huize.zhang@monash.edu
  - name: Dianne Cook
    affiliation: Monash University 
    address: Department of Econometrics and Business Statistics
    email:  dicook@monash.edu
  - name: Ursula Laa
    affiliation: University of Natural Resources and Life Sciences
    address: Institute of Statistics
    email:  ursula.laa@boku.ac.at  
  - name: Nicolas Langrené
    affiliation: CSIRO Data61 
    address: 34 Village Street, Docklands VIC 3008 Australia
    email: nicolas.langrene@csiro.au
  - name: Patricia Menéndez
    affiliation: Monash University 
    address: Department of Econometrics and Business Statistics
    email:  patricia.menendez@monash.edu 
abstract: |
  Guided tour searches for interesting low-dimensional views of high-dimensional data via optimising a projection pursuit index function. The first paper of projection pursuit by @friedman1974projection stated that "the technique used for maximising the projection index strongly influences both the statistical and the computational aspects of the procedure." While much work has been done in proposing  indices in the literature, less has been done on evaluating the performance of the optimisers. In this paper, we implement a data collection object in the optimisation of projection pursuit guided tour and introduce visual diagnostics based on the data object collected. These diagnostics and workflow can be applied to a broad class of optimisers, to assess their performance. An R package, \pkg{ferrn}, has been created to implement the diagnostics.

@@

```{r }
#spelling::spell_check_files(here::here("paper/zhang-cook-laa-langrene-menendez.Rmd"), lang = "en-GB", ignore = spelling::get_wordlist())
```

@@

Let $\mathbf{X}_{n \times p}$ be the data matrix with $n$ observations in $p$ dimensions. A $d$-dimensional projection is a linear transformation from $\mathbb{R}^p$ into $\mathbb{R}^d$ defined as $\mathbf{Y} = \mathbf{X} \cdot \mathbf{A}$, where $\mathbf{Y}_{n \times d}$ is the projected data and $\mathbf{A}_{p\times d}$ is the projection basis matrix. We define $f: \mathbb{R}^{n \times d} \mapsto \mathbb{R}$ to be an index function that maps the projected data $\mathbf{Y}$ onto a scalar value. This is commonly known as the projection pursuit index function, or just index function, and is used to measure the "interestingness" of a given projection. In projection pursuit, higher index values are associated with interesting projections that depart from normality [@diaconis1984asymptotics; @huber1985projection], e.g. Legendre index [@friedman1974projection], Hermite index [@hall1989polynomial] and natural Hermite index [@cook1993projection] for measuring the distance between the density of projected data and a standard normal distribution, LDA index [@lee2005projection] and PDA index [@lee2010projection] for finding clustering structure in the data, and skewness-based index [@Loperfido2018] and kurtosis-based index [@Loperfido2020] for outlier detection. As a general visualisation method, tour produces animations of high dimensional data via rotations of low-dimensional planes. There are different tour implementations depending on how the high dimensional space is investigated: grand tour [@cook2008grand] selects the planes randomly to provide a general overview;  manual tour [@cook1997manual] gradually phases in and out one variable to understand the contribution of that variable in the projection. Guided tour, the main interest of this paper, chooses the planes with the aid of projection pursuit, to gradually reveal the most interesting projection. Given a random start, projection pursuit iteratively finds bases with higher index values and the guided tour constructs a geodesic interpolation between these planes to form a tour path. Figure \ref{fig:tour-path} shows a sketch of the tour path where the blue frames are produced by the projection pursuit optimisation and the white frames are interpolations  between the blue frames. Mathematical details of the geodesic interpolation can be found in @buja2005computational. The aforementioned tour method has been implemented in the R package \CRANpkg{tourr} [@tourr].

### 2021-02-18 paper/zhang-cook-laa-langrene-menendez.tex
better examples of projection pursuit indices

\abstract{%
Guided tour searches for interesting low-dimensional views of
high-dimensional data via optimising a projection pursuit index
function. The first paper of projection pursuit by
\citet{friedman1974projection} stated that ``the technique used for
maximising the projection index strongly influences both the statistical
and the computational aspects of the procedure.'' While much work has
been done in proposing indices in the literature, less has been done on
evaluating the performance of the optimisers. In this paper, we
implement a data collection object in the optimisation of projeciton
pursuit guided tour and introduce visual diagnostics based on the data
object collected. These diagnostics and workflows can be applied to a
broad class of optimisers, to assess their performance. An R package,
\pkg{ferrn}, has been created to implement the diagnostics.

@@

Let \(\mathbf{X}_{n \times p}\) be the data matrix with \(n\)
observations in \(p\) dimensions. A \(d\)-dimensional projection is a
linear transformation from \(\mathbb{R}^p\) into \(\mathbb{R}^d\)
defined as \(\mathbf{Y} = \mathbf{X} \cdot \mathbf{A}\), where
\(\mathbf{Y}_{n \times d}\) is the projected data and
\(\mathbf{A}_{p\times d}\) is the projection basis matrix. We define
\(f: \mathbb{R}^{n \times d} \mapsto \mathbb{R}\) to be an index
function that maps the projected data \(\mathbf{Y}\) onto a scalar
value. This is commonly known as the projection pursuit index function,
or just index function, and is used to measure the ``interestingness''
of a given projection. In the context of projection pursuit,
interestingness means departure from normality
\citep{diaconis1984asymptotics, huber1985projection}.
\textcolor{red}{only include a couple of indices, using e.g.}A number of
index functions have been proposed in the literature to detect different
data structures
\citep{friedman1974projection, hall1989polynomial, cook1993projection, posse1995projection}
and to classify groups \citep{lee2005projection, lee2010projection}.

As a general visualisation method, tour produces animations of high
dimensional data via rotations of low-dimensional planes. There are
different tour implementations depending on how the high dimensional
space is investigated: grand tour \citep{cook2008grand} selects the
planes randomly to provide a general overview; manual tour
\citep{cook1997manual} gradually phases in and out one variable to
understand the contribution of that variable in the projection. Guided
tour, the main interest of this paper, chooses the planes with the aid
of projection pursuit, to gradually reveal the most interesting
projection. Given a random start, projection pursuit iteratively finds
bases with higher index values and the guided tour constructs a geodesic
interpolation between these planes to form a tour path. Figure
\ref{fig:tour-path} shows a sketch of the tour path where the blue
frames are produced by the projection pursuit optimisation and the white
frames are interpolations between the blue frames. Mathematical details
of the geodesic interpolation can be found in
\citet{buja2005computational}. The aforementioned tour method has been
implemented in the R package \CRANpkg{tourr} \citep{tourr}.

>>>>

\abstract{%
Guided tour searches for interesting low-dimensional views of
high-dimensional data via optimising a projection pursuit index
function. The first paper of projection pursuit by
\citet{friedman1974projection} stated that ``the technique used for
maximising the projection index strongly influences both the statistical
and the computational aspects of the procedure.'' While much work has
been done in proposing indices in the literature, less has been done on
evaluating the performance of the optimisers. In this paper, we
implement a data collection object in the optimisation of projection
pursuit guided tour and introduce visual diagnostics based on the data
object collected. These diagnostics and workflow can be applied to a
broad class of optimisers, to assess their performance. An R package,
\pkg{ferrn}, has been created to implement the diagnostics.

@@

Let \(\mathbf{X}_{n \times p}\) be the data matrix with \(n\)
observations in \(p\) dimensions. A \(d\)-dimensional projection is a
linear transformation from \(\mathbb{R}^p\) into \(\mathbb{R}^d\)
defined as \(\mathbf{Y} = \mathbf{X} \cdot \mathbf{A}\), where
\(\mathbf{Y}_{n \times d}\) is the projected data and
\(\mathbf{A}_{p\times d}\) is the projection basis matrix. We define
\(f: \mathbb{R}^{n \times d} \mapsto \mathbb{R}\) to be an index
function that maps the projected data \(\mathbf{Y}\) onto a scalar
value. This is commonly known as the projection pursuit index function,
or just index function, and is used to measure the ``interestingness''
of a given projection. In projection pursuit, higher index values are
associated with interesting projections that depart from normality
\citep{diaconis1984asymptotics, huber1985projection}, e.g.~Legendre
index \citep{friedman1974projection}, Hermite index
\citep{hall1989polynomial} and natural Hermite index
\citep{cook1993projection} for measuring the distance between the
density of projected data and a standard normal distribution, LDA index
\citep{lee2005projection} and PDA index \citep{lee2010projection} for
finding clustering structure in the data, and skewness-based index
\citep{Loperfido2018} and kurtosis-based index \citep{Loperfido2020} for
outlier detection. As a general visualisation method, tour produces
animations of high dimensional data via rotations of low-dimensional
planes. There are different tour implementations depending on how the
high dimensional space is investigated: grand tour \citep{cook2008grand}
selects the planes randomly to provide a general overview; manual tour
\citep{cook1997manual} gradually phases in and out one variable to
understand the contribution of that variable in the projection. Guided
tour, the main interest of this paper, chooses the planes with the aid
of projection pursuit, to gradually reveal the most interesting
projection. Given a random start, projection pursuit iteratively finds
bases with higher index values and the guided tour constructs a geodesic
interpolation between these planes to form a tour path. Figure
\ref{fig:tour-path} shows a sketch of the tour path where the blue
frames are produced by the projection pursuit optimisation and the white
frames are interpolations between the blue frames. Mathematical details
of the geodesic interpolation can be found in
\citet{buja2005computational}. The aforementioned tour method has been
implemented in the R package \CRANpkg{tourr} \citep{tourr}.
